Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-04-29 appendix.tex

\section{Full governing equation for vertically isothermal disks in
  the vertically global shearing box}\label{adia_improve}
Here we explain the derivation of Eq. \ref{nearly_iso_explicit} used in our 
analytical discussion in \S\ref{analytical}. We shall need the
vertical derivative of the epicyclic frequency $\kappa$. In the
vertically isothermal global
disk with $\Gamma=1$ we have 
\begin{align}\label{dkappa2}
  \frac{\p\kappa^2}{\p z} = 4 \frac{\p\Omega^2}{\p z} + r\frac{\p}{\p
    r}\frac{\p\Omega^2}{\p z} = -
  \frac{\p\ln\rho}{\p z}\frac{qc_s^2}{r^2} \left(
    \frac{3z^2}{r^2+z^2}-1\right) \equiv - \frac{\p\ln\rho}{\p z}\frac{qc_s^2}{r^2}F(r,z).
\end{align}
% For the local problem we may then write
% \begin{align}
%   \frac{dD}{dz}  = - \frac{d\ln\rho}{dz}\frac{qc_s^2}{r^2}F(z),
% \end{align}
The function $F$ increases monotonically from $F=-1$ at $z=0$ to $F\to2$
as $|z|\to\infty $, so $|F|=O(1)$.

In the vertically global shearing box (VGSB) approximation, we set
$\zeta=0$ in Eq. \ref{lin_mass} --- \ref{lin_energy}. We then eliminate
the horizonatal velocity perturbations ($\delta v_x,\, \delta v_y$) to
obtain  
%we can eliminate variables 
%to obtain a pair or ordinary differential
%equations for $W,Q$ as 
\begin{align}
  & \ii\upsilon\delta v_z = \frac{dW}{dz} + \frac{d\ln{\rho}}{d z}\left(W-Q\right),\label{ode_vz} \\
  &\frac{\upsilon^2}{c_s^2}Q + \frac{\upsilon^2k_x^2}{D}W =\ii\upsilon \left(\frac{\ii
      k_x r}{D}\frac{d\Omega^2}{dz} -
    \frac{d\ln{\rho}}{dz}\right)\delta v_z - \ii\upsilon \frac{d\delta v_z}{dz},
  % \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right] 
  % \notag\\
  % &- \frac{d^2W}{dz^2} - \frac{d\ln{\rho}}{dz}\left(\frac{dW}{dz} -
  %   \frac{dQ}{dz}\right) - \frac{d^2\ln{\rho}}{dz^2}\left(W-Q\right),
  \label{ode_w}\\
%  &\sigma^2W - \frac{\gamma}{\Gamma}\sigma^2 Q +
%  \frac{\ii\sigma}{t_c}\left(W-\frac{Q}{\Gamma}\right) 
%  =\ii \sigma c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma} -
%    1\right)\delta v_z,  
  &\upsilon^2W - \gamma\upsilon^2 Q +
  \frac{\ii\upsilon}{t_c}\left(W-Q\right) 
  =\ii \upsilon c_s^2\frac{d\ln{\rho}}{dz}\left(\gamma -
    1\right)\delta v_z,
  % 
  % \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right],
  \label{ode_Q} 
\end{align}
(note that $\Gamma=1$ has been used) where
\begin{align}
  D \equiv \kappa^2 - \upsilon^2,
\end{align} 
and we have replaced the notation $\p_z$ by $d/dz$ since we are
considering solutions at a fixed radius.

@@

We now eliminate $W,\, Q$ from Eq. \ref{ode_vz} --- \ref{ode_Q} in
favor of $\delta v_z$, making use of Eq. \ref{dkappa2} to evaluate
$dD/dz$, to obtain

\begin{align}\label{full_ode}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left[1 + \frac{\ii k_x c_s^2
      q}{Dr} - \frac{k_x^2c_s^2}{\left(k_x^2c_s^2 + \chi
        D\right)}\frac{qc_s^2F}{Dr^2}\right]\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left\{\upsilon^2\left(\frac{k_x^2}{D} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{Dr}\right)\frac{d^2\ln\rho}{dz^2} -
    \frac{c_s^2}{D}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left[\left(1-\chi\right) +
     \frac{\chi}{\left(k_x^2c_s^2 + \chi D\right)}\frac{qc_s^2 F}{r^2}\right] 
   \right\}\delta v_z,
\end{align}
where we recall $\chi = \left(1-\ii\upsilon t_c\right)/\left(1-\ii\upsilon
t_c \gamma\right)$. Making the low frequency and Keplerian
approximations is equivalent to the replacement $D\to
\OmK^2$. This gives, in terms of dimensionless variables,  
\begin{align}
  &\delta v_z ^{\prime\prime} + \left[1 + \ii h q\hat{k} -
    \frac{ \hat{k}^2}{
      \left(\hat{k}^2+\chi\right)}q h^2F\right]\ln\rho^{\prime}\delta v_z^\prime +
  \left\{\left(\chi + \ii h q
      \hat{k}\right)\ln\rho^{\prime\prime} - \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii h
      q\hat{k}\right)\left[1 - \chi +
      \frac{\chi}{\left(\hat{k}^2+\chi\right)}qh^2F\right]\right\}\delta v_z \notag\\&=
  -\hat{\upsilon}^2\left(\hat{k}^2+\chi\right)\delta v_z\label{adia_diso3}.
\end{align} 
Since we are interested in thin disks with $h \ll 1$ and $|q| =
O(1)$, we neglect the terms proportional to $h^2$, giving 
\begin{align}
  0= \delta v_z ^{\prime\prime} + \left(1 + \ii h
     q\hat{k}\right)\ln\rho^{\prime}\delta v_z^\prime 
   +
   \left[\hat{\upsilon}^2\left(\hat{k}^2+\chi\right) +
     \left(  \chi + \ii h q\hat{k}\right)\ln\rho^{\prime\prime}
     - \ln\rho^{\prime
       2}\left(\khat^2 -
       \ii h
       q\hat{k}\right)\left(1 - \chi\right)\right]\delta v_z,\label{vertiso_gov_nondim}
 \end{align}
 which is used in the discussion below.  Eq. \ref{nearly_iso_explicit}, as
 used in our analytical discussion, is obtained by inserting the 
 thin-disk expression for the density field. We comment that the simpler way
 of obtaining Eq. \ref{vertiso_gov_nondim} is to neglect the vertical 
 dependence of $D$ by setting $D\to\OmK^2$ in Eq. \ref{ode_w} before eliminating
 variables in favor of $\delta v_z$.

>>>>

\section{Derivation of the approximate equations}\label{adia_improve}
Here we detail the derivation of Eq. \ref{nearly_iso_explicit} used in the 
analytical derivations of \S\ref{analytical}.  Starting with Eqs.\ \ref{lin_mass} --- \ref{lin_energy},
we set $\Gamma = 1$ and $\zeta = 0$ for the vertically isothermal limit and the fully-radially-local 
approximation, respectively.  We  eliminate
the horizontal velocity perturbations ($\delta v_x,\, \delta v_y$) to
obtain  
%we can eliminate variables 
%to obtain a pair or ordinary differential
%equations for $W,Q$ as 
\begin{subequations}\label{A1}\begin{align}
  & \ii\upsilon\delta v_z = \frac{dW}{dz} + \frac{d\ln{\rho}}{d z}\left(W-Q\right),\label{ode_vz} \\
  &\frac{\upsilon^2}{c_s^2}Q + \frac{\upsilon^2k_x^2}{D}W =\ii\upsilon \left(\frac{\ii
      k_x r}{D}\frac{d\Omega^2}{dz} -
    \frac{d\ln{\rho}}{dz}\right)\delta v_z - \ii\upsilon \frac{d\delta v_z}{dz},
  % \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right] 
  % \notag\\
  % &- \frac{d^2W}{dz^2} - \frac{d\ln{\rho}}{dz}\left(\frac{dW}{dz} -
  %   \frac{dQ}{dz}\right) - \frac{d^2\ln{\rho}}{dz^2}\left(W-Q\right),
  \label{ode_w}\\
%  &\sigma^2W - \frac{\gamma}{\Gamma}\sigma^2 Q +
%  \frac{\ii\sigma}{t_c}\left(W-\frac{Q}{\Gamma}\right) 
%  =\ii \sigma c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma} -
%    1\right)\delta v_z,  
  &\upsilon^2W - \gamma\upsilon^2 Q +
  \frac{\ii\upsilon}{t_c}\left(W-Q\right) 
  =\ii \upsilon c_s^2\frac{d\ln{\rho}}{dz}\left(\gamma -
    1\right)\delta v_z,
  % 
  % \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right],
  \label{ode_Q} 
\end{align}\end{subequations}
where
\begin{align}
  D \equiv \kappa^2 - \upsilon^2,
\end{align} 
and we have replaced $\p/\p z$ by $d/dz$ for the fully radially local treatment.

Reduction to a single ODE requires vertical derivatives.  At this point we could use the 
low frequency and Keplerian approximations to set $D \rightarrow \kappa^2 \rightarrow \OmK^2$,
 and we would wind up with Eq. \ref{nearly_iso_explicit} more directly.
However to demonstrate that the order of approximation is irrelevant, we will retain
the vertical derivatives of the epicyclic frequency $\kappa$,
\begin{align}\label{dkappa2}
  \frac{d\kappa^2}{d z} = 4 \frac{\p\Omega^2}{\p z} + r\frac{\p}{\p
    r}\frac{\p\Omega^2}{\p z} = -
  \frac{d\ln\rho}{d z}\frac{qc_s^2}{r^2} \left(
    \frac{3z^2}{r^2+z^2}-1\right) \equiv - \frac{d\ln\rho}{d z}\frac{qc_s^2}{r^2}F(r,z).
\end{align}
% For the local problem we may then write
% \begin{align}
%   \frac{dD}{dz}  = - \frac{d\ln\rho}{dz}\frac{qc_s^2}{r^2}F(z),
% \end{align}
The function $F$ increases monotonically from $F=-1$ at $z=0$ to $F\to2$
as $|z|\to\infty $, so $|F|=O(1)$.

@@

We now eliminate $W,\, Q$ from Eqs.\ \ref{A1}, using  Eqs. \ref{chi} and \ref{dkappa2} to obtain  
\begin{align}\label{full_ode}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left[1 + \frac{\ii k_x c_s^2
      q}{Dr} - \frac{k_x^2c_s^2}{\left(k_x^2c_s^2 + \chi
        D\right)}\frac{qc_s^2F}{Dr^2}\right]\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left\{\upsilon^2\left(\frac{k_x^2}{D} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{Dr}\right)\frac{d^2\ln\rho}{dz^2} -
    \frac{c_s^2}{D}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left[\left(1-\chi\right) +
     \frac{\chi}{\left(k_x^2c_s^2 + \chi D\right)}\frac{qc_s^2 F}{r^2}\right] 
   \right\}\delta v_z.
\end{align}

Now we make the low frequency and Keplerian
approximations, setting $D\to
\OmK^2$, to give  
\begin{align}
  &\delta v_z ^{\prime\prime} + \left[1 + \ii h q\hat{k} -
    \frac{ \hat{k}^2}{
      \left(\hat{k}^2+\chi\right)}q h^2F\right]\ln\rho^{\prime}\delta v_z^\prime +
  \left\{\left(\chi + \ii h q
      \hat{k}\right)\ln\rho^{\prime\prime} - \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii h
      q\hat{k}\right)\left[1 - \chi +
      \frac{\chi}{\left(\hat{k}^2+\chi\right)}qh^2F\right]\right\}\delta v_z \notag\\&=
  -\hat{\upsilon}^2\left(\hat{k}^2+\chi\right)\delta v_z\label{adia_diso3},
\end{align} 
in terms of dimensionless variables of Eq.\ \ref{nondim}.

Retaining terms to first order in the disk aspect ratio, $h \ll 1$, gives
\begin{align}
  0= \delta v_z ^{\prime\prime} + \left(1 + \ii h
     q\hat{k}\right)\ln\rho^{\prime}\delta v_z^\prime 
   +
   \left[\hat{\upsilon}^2\left(\hat{k}^2+\chi\right) +
     \left(  \chi + \ii h q\hat{k}\right)\ln\rho^{\prime\prime}
     - \ln\rho^{\prime
       2}\left(\khat^2 -
       \ii h
       q\hat{k}\right)\left(1 - \chi\right)\right]\delta v_z.\label{vertiso_gov_nondim}
 \end{align}
Applying Eq.\ \ref{rhoisothin} for the density field gives the desired result, Eq. \ref{nearly_iso_explicit}.

We can establish a correspondence between our  Eq. \ref{nearly_iso_explicit} and
 Eq. 41 in \cite{lubow93}, which describes adiabatic axisymmetric waves in
a vertically isothermal disk without vertical shear.   Accounting for the required change of variables, 
the correspondence is exact after setting $q=0$ (no vertical shear) and
$\chi=1/\gamma$ (adiabatic flow) in our Eq. \ref{nearly_iso_explicit},
and applying the approximations in our \S \ref{sec:simplified} to \citeauthor{lubow93}.

%We comment that the simpler way
 %of obtaining Eq. \ref{vertiso_gov_nondim} is to neglect the vertical 
 %dependence of $D$ by setting $D\to\OmK^2$ in Eq. \ref{ode_w} before eliminating
 %variables in favor of $\delta v_z$.

### 2015-04-29 linear_adia.tex

\section{Analytic results for vertically isothermal disks}\label{analytical}

\subsection{Isothermal perturbations}\label{iso_pert}
Isothermal perturbations correspond to $\beta = 0$ or $\gamma=1$ so 
that $\chi = 1$. This case has been investigated by \citetalias{nelson13},
\citetalias{mcnally14} and \citetalias{barker15}.

In Appendix \ref{iso_discuss}, we collect some useful results for
isothermal  perturbations. This includes a necessary condition for
instability (there must be vertical shear) and an upper bound to the
instability growth rate (it is limited by the maximum vertical shear
rate in the disk). Neither of these actually require the thin-disk
form of the density field described above.  
We also develop solutions in the thin-disk
limit, but without assuming low frequency from the onset. 
The methods used in Appendix \ref{iso_discuss} and results obtained
therein motivates the solution procedure employed in the following
sections, and are provided for ease of reference. We refer the reader
to \citetalias{barker15} for a more detailed treatment of isothermal 
perturbations. %which also account for more general boundary
               %conditions than that considered below

\subsection{The effect of introducing a small but finite
  thermal relaxation time}\label{relax_pert}
We are interested in finite thermal relaxation
timescales $\beta > 0$, but it is instructive to
first ask  the more analytically tractable question: how 
do the eigenfrequencies and eigenfunctions change when we change
$\beta$ from zero to a small but finite value? For 
sufficiently small $\beta$ we expect the solution to only differ
slightly from a case with $\beta\equiv 0$. We thus perturb a solution
for $\beta=0$ to see the effect of finite cooling.

@@

The fundamental mode has been observed to dominate numerical simulations
\citep[\citetalias{nelson13}]{stoll14}, and are the ones we find to typically 
dominate in numerical calculations with increasing $\beta$
(\S\ref{therm_relax_eff}) for moderately large    
values of $\khat$, as well as in PPDs 
with a realistic estimate for thermal timescales
(\S\ref{application}). We will see later that consideration of low
order modes in fact provides a useful way to charactertize the effect
of increasing the thermal timescale on the VSI.

@@

\subsection{Explicit solutions and dispersion relation}\label{disp_relax}
We now solve Eq. \ref{nearly_iso_explicit} explicitly. We first write  
\begin{align}
  \delta v_z(\zhat) =
  g(\zhat)\exp\left(\frac{\lambda\zhat^2}{2}\right), \label{adia_ansatz}
\end{align}
where $\lambda$ is a constant to be chosen for convenience. Inserting
Eq. \ref{adia_ansatz} into Eq. \ref{nearly_iso_explicit} gives
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\lambda\right)g^\prime + \left(B +
    \lambda\right)g
  +\left(\lambda^2 - \lambda A - C\right)\zhat^2 g.
\end{align}
We choose $\lambda$ to make the coefficient of $\zhat^2g$
vanish, and impose the vertical kinetic energy density
$\rho|\delta v_z|^2\propto |g|^2 \exp{\left(\real\lambda -
    1/2\right)\zhat^2}$ to remain finite as $|\zhat|\to\infty$. 
Then assuming $g(\zhat)$ is a polynomial, we require  
\begin{align}
  \real\lambda < \frac{1}{2}, 
\end{align}
which amounts to choosing 
\begin{align}
  \lambda = \frac{1}{2}\left(A - \sqrt{A^2 + 4C}\right).  
\end{align}

@@

\subsection{Upper limit on the thermal relaxation timescale for the
  VSI}\label{iso_vsi_beta_crit}
Here we estimate the maximum thermal relaxation timescale 
$\beta$ that allows the VSI.  
Our starting point is Eq. \ref{relax_cond1}---\ref{relax_cond2} which
approximates the dispersion relation Eq. \ref{relax_disp} assuming marginal
stability and $\khat^2\gg 1$. We consider $M=0$ (the 
fundamental mode) or  $M=O(1)$ and low frequency $|\hat{\omega}|\ll
1$ such that $|\hat{\omega}\khat|$ is $O(1)$. We 
also assume $|q|$ and $\beta$ are $O(1)$ or smaller. For a thin disk
($h \ll 1$) we have 
\begin{align}
 0 = &M(M+1)\left(1-h^2q^2\khat^2\right) \notag\\ 
 &+ \left[1 +
    \gamma\beta^2\left(1-\gamma\right)(1+2M)^2\right]\left(\hat{\omega}\khat\right)^2
   -  \left(\hat{\omega}\khat\right)^4,\label{relax_cond_simp1}\\
   0= &2M(M+1) + \frac{\beta}{h q} (1-\gamma)(1+2M)^2
   \left(\hat{\omega}\khat\right) + \left(\hat{\omega}\khat\right)^2. \label{relax_cond_simp2}
\end{align}
(We explain this in Appendix \ref{disp_neut_limit}.)

>>>>

\section{Analytic results with finite cooling times}\label{analytical}

%\subsection{Isothermal perturbations}\label{iso_pert}
Previous analytic studies of the VSI have largely focused on isothermal
perturbations, with infinitely rapid cooling, as discussed in the introduction.  We further explore
this $\beta = 0$ limit in Appendix \ref{iso_discuss} both to further develop intuition for this 
idealized case and to establish a connection with previous 
works.  However our main interest is the effect of finite cooling times, which
we explore analytically in this section.

In \S\ref{relax_pert} we prove that even an infinitesimal increase in the cooling time,
starting from $\beta = 0$, slows the growth rate of the VSI.  In \S\ref{disp_relax} we 
derive exact solutions to the simplified VSI model developed above (Eq.\ \ref{nearly_iso_explicit}).
In \S\ref{iso_vsi_beta_crit} we reach our main result, the maximum cooling time above which 
VSI growth is suppressed.  We heuristically argue for this result in \S\ref{vsi_require} and numerically
verify it in \S\ref{numerical}.

%Isothermal perturbations correspond to $\beta = 0$ or $\gamma=1$ so 
%that $\chi = 1$. This case has been investigated by \citetalias{nelson13},
%\citetalias{mcnally14} and \citetalias{barker15}.  
%
%In Appendix \ref{iso_discuss}, we collect some useful results for
%isothermal  perturbations. This includes a necessary condition for
%instability (there must be vertical shear) and an upper bound to the
%instability growth rate (it is limited by the maximum vertical shear
%rate in the disk). Neither of these actually require the thin-disk
%form of the density field described above.  
%We also develop solutions in the thin-disk
%limit, but without assuming low frequency from the onset. 
%The methods used in Appendix \ref{iso_discuss} and results obtained
%therein motivates the solution procedure employed in the following
%sections, and are provided for ease of reference. We refer the reader
%to \citetalias{barker15} for a more detailed treatment of isothermal 
%perturbations. %which also account for more general boundary
%               %conditions than that considered below

\subsection{Introducing a small but finite
  cooling time}\label{relax_pert}
We are interested in finite thermal relaxation
timescales $\beta > 0$, but it is instructive to
first ask  the more analytically tractable question: how 
do the eigenfrequencies and eigenfunctions change when we change
$\beta$ from zero to a small but finite value? For 
sufficiently small $\beta$ we expect the solution to only differ
slightly from a case with $\beta\equiv 0$. We thus perturb a solution
for $\beta=0$ to see the effect of finite cooling.

@@

The fundamental mode has been observed to dominate numerical simulations
\citep[\citetalias{nelson13}]{stoll14}, and are the ones we find to typically 
dominate in numerical calculations with increasing $\beta$
(\S\ref{therm_relax_eff}) for moderately large    
values of $\khat$, as well as in PPDs 
with a realistic estimate for thermal timescales
(\S\ref{application}). We will see later that consideration of low
order modes in fact provides a useful way to characterize the effect
of increasing the thermal timescale on the VSI.

@@

\subsection{Explicit solutions and dispersion relation}\label{disp_relax}
We now solve Eq. \ref{nearly_iso_explicit} explicitly. We first write  
\begin{align}
  \delta v_z(\zhat) =
  g(\zhat)\exp\left(\frac{\lambda\zhat^2}{2}\right), \label{adia_ansatz}
\end{align}
where the constant $\lambda$ is a constant to be chosen for convenience. Inserting
Eq. \ref{adia_ansatz} into Eq. \ref{nearly_iso_explicit} gives
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\lambda\right)g^\prime + \left(B +
    \lambda\right)g
  +\left(\lambda^2 - \lambda A - C\right)\zhat^2 g.
\end{align}
We choose $\lambda$ to make the coefficient of $\zhat^2g$
vanish, and impose the vertical kinetic energy density
$\rho|\delta v_z|^2\propto |g|^2 \exp{\left(\real\lambda -
    1/2\right)\zhat^2}$ to remain finite as $|\zhat|\to\infty$. 
Then assuming $g(\zhat)$ is a polynomial, we require  
\begin{align}
  \real\lambda < \frac{1}{2}, 
\end{align}
which amounts to choosing 
\begin{align}
  \lambda = \frac{1}{2}\left(A - \sqrt{A^2 + 4C}\right).  
\end{align}

@@

\subsection{An upper limit to the cooling time}\label{iso_vsi_beta_crit}
Here we estimate the maximum thermal relaxation timescale 
$\beta$ that allows the VSI.  
Our starting point is Eq. \ref{relax_cond1}---\ref{relax_cond2} which
approximates the dispersion relation Eq. \ref{relax_disp} assuming marginal
stability and $\khat^2\gg 1$. We consider $M=0$ (the 
fundamental mode) or  $M=O(1)$ and low frequency $|\hat{\omega}|\ll
1$ such that $|\hat{\omega}\khat|$ is $O(1)$. We 
also assume $|q|$ and $\beta$ are $O(1)$ or smaller. For a thin disk
($h \ll 1$) we have 
\begin{align}
 0 = &M(M+1)\left(1-h^2q^2\khat^2\right) \notag\\ 
 &+ \left[1 +
    \gamma\beta^2\left(1-\gamma\right)(1+2M)^2\right]\left(\hat{\omega}\khat\right)^2
   -  \left(\hat{\omega}\khat\right)^4,\label{relax_cond_simp1}\\
   0= &2M(M+1) + \frac{\beta}{h q} (1-\gamma)(1+2M)^2
   \left(\hat{\omega}\khat\right) + \left(\hat{\omega}\khat\right)^2, \label{relax_cond_simp2}
\end{align}
as explained in  Appendix \ref{disp_neut_limit}.

### 2015-04-29 linear_iso.tex

In preparation for a numerical study of the above equations, we will 
first give a fully analytic discussion for vertically isothermal
disks, for which $\Gamma=1$ and $c_s$ is vertically constant. To this end, 
we introduce the following simplifications:
%In Appendix \ref{adia_improve} we derive  a with the following simplifications:

@@

\item  We adopt the  `vertically global shearing box' (VGSB) formalism 
  developed by \citetalias{mcnally14}, in which explicit dependencies on the
  background radial disk structure in 
  Eq. \ref{lin_mass}---\ref{lin_energy} are neglected by setting
  $\zeta=0$. However, the global radial temperature 
  gradient in the expression for the 
  vertical shear (Eq. \ref{vertical_shear}) is retained.

\item We consider inertial waves with eigenfrequencies such that
  $|\upsilon^2|\ll \kappa^2,\,\Omega^2$ \citepalias{barker15}. We call
  this the low-frequency approximation.

\item We are mainly interested in the effect of vertical shear, 
%  i.e. $q\neq
%  0$ for vertically isothermal disks. 
  which is already captured
  explicitly in the linearized equations through the term
  $\p_z\Omega$. We thus ignore the vertical dependence of $\Omega$ and
  $\kappa$ elsewhere, and replace them with
  $\OmK$.  We call this the Keplerian approximation.

\item We assume thin disks for which $h\ll 1$, and write 
  $\rho/\rho_0\simeq  \exp{(-z^2/2H^2)}$.

@@

 In terms of the dimensionless variables
\begin{align}
  \hat{z} = z/H,\quad \hat{k}=k_xH, \quad \hat{\upsilon} =\hat{\omega} +
  \ii\hat{\sigma}= \upsilon/\OmK,
\end{align}
where $\hat{\omega}$ and $\hat{\sigma}$ are real, the above
approximations lead to a single second order ODE, 
% Eq. \ref{vertiso_gov} becomes 
% \begin{align}
%   0=& \delta v_z ^{\prime\prime} + \left(1 + \ii h
%     q\hat{k}\right)\ln\rho^{\prime}\delta v_z^\prime \notag\\
%   &+
%   \left[\hat{\upsilon}^2\left(\hat{k}^2+\chi\right) +
%     \left(  \chi + \ii h q\hat{k}\right)\ln\rho^{\prime\prime}\right.
%   \notag\\
%   &\left.- \ln\rho^{\prime
%       2}\left(\khat^2 -
%       \ii h
%       q\hat{k}\right)\left(1 - \chi\right)\right]\delta v_z,\label{vertiso_gov_nondim}
% \end{align}
% where $^\prime$ denotes $d/d\zhat$.
%\subsection{Thin-disk approximation}\label{analytic_relax}
%The governing equation, Eq. \ref{vertiso_gov_nondim}, is then   
\begin{align}
  \delta v_z^{\prime\prime} - z A\delta v_z^\prime +
  (B - C\zhat^2)\delta v_z = 0,\label{nearly_iso_explicit}
\end{align}
where $^\prime$ denotes $d/d\zhat$ and 
\begin{align}
  &A \equiv 1 + \ii h q \hat{k},\\
  &B \equiv \hat{\upsilon}^2\left(\chi + \hat{k}^2\right) -
  \left(\chi + \ii h q \hat{k}\right),\\
  &C \equiv \left(1-\chi\right)\left(\hat{k}^2 - \ii
    h q\hat{k}\right), 
\end{align}
where
\begin{align}
\chi = \frac{1-\ii\hat{\upsilon}\beta}{1-\ii\hat{\upsilon}\beta\gamma}.
\end{align}
The derivation of Eq. \ref{nearly_iso_explicit}  is outlined in
Appendix \ref{adia_improve}.

We remark that Eq. \ref{nearly_iso_explicit} is in the same 
form as Eq. 41 in \cite{lubow93}, which describe adiabatic axisymmetric waves in
a vertically isothermal disk without vertical shear. The two equations
become identical after  setting $q=0$ (no vertical shear) and
$\chi=1/\gamma$ (adiabatic flow) in our Eq. \ref{nearly_iso_explicit},
and making the low-frequency, Keplerian approximation in
\citeauthor{lubow93} after a change of variables.

 In Appendix \ref{global_corr}  we discuss limitations of using
  the VGSB, where we show that this approximation is valid provided we 
  consider fast cooling times
  ($\beta\lesssim 1$) and perturbations with small radial lengthscales
  ($\khat\gtrsim 1$), as appropriate for the VSI.

>>>>

Our simplified model starts with  Eqs.\ \ref{lin_all} 
and makes the following additional approximations:

@@

 \item We set $\Gamma = 1$, focusing on vertically isothermal disks.

\item  We set $\zeta = 0$,  dropping terms with an explicit dependence on the 
radial structure of the equilibrium disk.  Vertical shear, which implicitly depends on the radial
temperature gradient, is retained.  This fully-radially-local approximation is also made in the
 `vertically global shearing box' (VGSB) of \citetalias{mcnally14}.

\item We make the low frequency approximation, $|\upsilon^2|\ll \kappa^2,\,\Omega^2$,
keeping only the lowest order terms in $|\upsilon|/\kappa$.  Similar to the 
incompressible \citepalias{goldreich67} or anelastic \citepalias{barker15} approximations,
the low frequency approximation filters p-modes in favor of g-modes, i.e.\ inertial
gravity waves \citep{lubow93}.

\item We make the Keplerian approximation, setting $\Omega \rightarrow \OmK$ 
and $\kappa \rightarrow \OmK$, but retaining the vertical dependence 
in the crucial vertical shear term,
  $\p_z\Omega$.

\item We assume thin disk with $h\ll 1$ and take the $z \ll r$ limit to 
  obtain Eq.\ \ref{rhoisothin} for the equilibrium $\rho(z)$.

@@

 In terms of the dimensionless variables
\begin{align}\label{nondim}
  \hat{z} = z/H,\quad \hat{k}=k_xH, \quad \hat{\upsilon} =\hat{\omega} +
  \ii\hat{\sigma}= \upsilon/\OmK,
\end{align}
where $\hat{\omega}$ and $\hat{\sigma}$ are real, the above
approximations lead to a single second order ODE, 
% Eq. \ref{vertiso_gov} becomes 
% \begin{align}
%   0=& \delta v_z ^{\prime\prime} + \left(1 + \ii h
%     q\hat{k}\right)\ln\rho^{\prime}\delta v_z^\prime \notag\\
%   &+
%   \left[\hat{\upsilon}^2\left(\hat{k}^2+\chi\right) +
%     \left(  \chi + \ii h q\hat{k}\right)\ln\rho^{\prime\prime}\right.
%   \notag\\
%   &\left.- \ln\rho^{\prime
%       2}\left(\khat^2 -
%       \ii h
%       q\hat{k}\right)\left(1 - \chi\right)\right]\delta v_z,\label{vertiso_gov_nondim}
% \end{align}
% where $^\prime$ denotes $d/d\zhat$.
%\subsection{Thin-disk approximation}\label{analytic_relax}
%The governing equation, Eq. \ref{vertiso_gov_nondim}, is then   
\begin{align}
  \delta v_z^{\prime\prime} - z A\delta v_z^\prime +
  (B - C\zhat^2)\delta v_z = 0,\label{nearly_iso_explicit}
\end{align}
where $^\prime$ denotes $d/d\zhat$ and 
\begin{subequations}\begin{align}
  &A \equiv 1 + \ii h q \hat{k},\\
  &B \equiv \hat{\upsilon}^2\left(\chi + \hat{k}^2\right) -
  \left(\chi + \ii h q \hat{k}\right),\\
  &C \equiv \left(1-\chi\right)\left(\hat{k}^2 - \ii
    h q\hat{k}\right), 
\end{align}\end{subequations}
where
\begin{align}\label{chi}
\chi = \frac{1-\ii\hat{\upsilon}\beta}{1-\ii\hat{\upsilon}\beta\gamma}.
\end{align}
The derivation of Eq. \ref{nearly_iso_explicit}  is detailed in
Appendix \ref{adia_improve}.

%We remark that Eq. \ref{nearly_iso_explicit} is in the same 
%form as Eq. 41 in \cite{lubow93}, which describe adiabatic axisymmetric waves in
%a vertically isothermal disk without vertical shear. The two equations
%become identical after  setting $q=0$ (no vertical shear) and
%$\chi=1/\gamma$ (adiabatic flow) in our Eq. \ref{nearly_iso_explicit},
%and making the low-frequency, Keplerian approximation in
%\citeauthor{lubow93} after a change of variables.

 In Appendix \ref{global_corr}  we discuss the limits of these approximations
 and show that they are valid in the parameter space of interest for the VSI.
In particular we point out that the fully-radially-local approximation is only valid for
short cooling times ($\beta \lesssim 1$).
